\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-16 metatheory.tex
Save

% Motivation of using QIIT
We start with the formulation of the syntax in the declarative style, which is based on the
formalization of a predicative \textit{Martin-Lof Type Theory} (MLTT)
given by~\citet{coquand2018canonicity}. 
% Thus we will omit a great deal of details about MLTT here
\YZ{What details are omitted?} 
We will mainly focus on the new facility that we introduced

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic an extensional type theoretical framework (with Agda-like syntax), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Our work can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We use as the formulation the intrinsically-typed debruijn syntax in this type-theoretical QIIT instead of a more conventional and untyped representation, because (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a proof because proof checking then is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real theorem prover so that the reader can have an idea about how mechanized formalization can be carried out (5) We can still use named-variables for intuition and example illustration (6) using QIIT as syntax encoding can avoid the reader from the concept of \textit{presupposition}~\cite{gratzer-sterling-birkedal-2019}---these \textit{presupposition} can be understood as \textit{implicit parameters} of the declaration of the judgemental rules in QIIT.

>>>>

% Motivation of using QIIT
We start with the formulation of the syntax in the declarative style, which is based on the
formalization of a predicative \textit{Martin-Lof Type Theory} (MLTT)~\cite{martin1982constructive}
given by~\citet{coquand2018canonicity}. 
% Thus we will omit a great deal of details about MLTT here
\YZ{What details are omitted?} 
We will mainly focus on the new facility that we introduced

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic an extensional type theoretical framework (with Agda-like syntax), 
together with \textit{Quotient Inductive Inductive Type} (QIIT).

\textit{Quotient Inductive Inductive Type}~(QIIT) is an extension of \textit{inductive-inductive type} (IIT)~\cite{nordvall2010inductive}. IIT can be roughly considered as several inductive type mutually defined, in a more expressible way than mutual-inductive type. QIIT is allowing to have \textit{quotient}, in other words, equality results as part of the constructors. For example, in \\ \mintinline{Agda}{data N2: Set where O : N2 | S : N2 → N2 | B : S (S x) ≡ x }, the "B" here is a valid (equality) constructor, and "N2" is a set of two elements in our extensional setting.

Our formulation can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We use as the formulation the intrinsically-typed debruijn syntax in this type-theoretical QIIT instead of a more conventional and untyped representation, because (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a pen-and-paper proof now because proof checking then is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real mechanized one so that the reader can have an idea about how mechanization can be carried out (5) We can still use named-variables for intuition and example illustration (6) using QIIT as syntax encoding can avoid the reader from the concept of \textit{presupposition}~\cite{gratzer-sterling-birkedal-2019}---these \textit{presupposition} can be understood as \textit{implicit parameters} of the declaration of the judgemental rules in QIIT.

### 2022-07-16 syntax.tex
Save

\subsubsection{\textbf{Warm Up}}
\textit{Quotient Inductive Inductive Type}~(QIIT) is an extension of \textit{inductive-inductive type} (IIT)~\cite{nordvall2010inductive}. IIT can be roughly considered as several inductive type mutually defined, in a more expressible way than mutual-inductive type. QIIT is allowing to have \textit{quotient}, in other words, equality results as part of the constructors. For example, in \mintinline{Agda}{data N2: Set where O : N2 | S : N2 → N2 | B : S (S x) ≡ x }, "B" will be a valid (equality) constructor, and "N2" is a set of two elements in our extensional setting.

We highlight some specialities of declarative style in QIIT that differ from a
conventional setting: 
(1) This declarative formulation of type theory does not
have an operational semantics but only judgemental equality---$\beta$-reduction is now $\beta$-equality.\footnote{Later we will prove
canonicity which can imply a certain level of computational ability of
our theory} 
(2) We directly use the equality from QIIT to represent judgemental equality in our syntax, relieve from us the burden to explicit mentioning equivalence closure; however, this also mean that, \textit{every piece of the syntax data}, including all kinds of judgements (QIIT type) and their derivations will have equality information upon them  
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
% (3) We are dealing with quotient data, so we are actually manipulating
% \textit{equivalence classes of terms} instead of one representative (term) from the equivalence class.
(3) Instead of using meta-level substitution, we use explicit
substitution and de Bruijn indices in our intrinsic formulation. This is
also called substitution calculus in the literature and is favoured due
to its closedness to categorical semantic. We will go through these details one-by-one.

\underline{\textit{\textbf{Judgements and Universe.}}}
% Judgements in QIIT
We formalize our language as a deep embedding in a type theory with QIIT types. Every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms from (vanilla) MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing} and thus a predicative MLTT, but for readability we
omit the universe levels.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the (pseudo) Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their (pseudo-)Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just as
a well-formed QIIT type is not necessarily inhabited. 
These well-formedness rules are direct translations of the \textit{well-typedness} of QIIT.

@@

% Example : how judgemental equality is used for the universe
We review the concept of ``Universe'' and at the same time showing how equality is representing semantic. In dependent type setting, function now can return type, of which the type is ``type of type'', i.e., \textit{the universe} $\cU$. We follow the style of \textit{type universe à la Tarski}~\cite{hofmann1997syntax} and
distinguish types from their \textit{codes or names} as showin in \ruleref{Codes/Get Type}: given a \textit{code} (actually a term) $\goodTerm{\Gamma}{T}{\cU}$ we can retrieve the corresponding type
$\goodType{\Gamma}{El~T}{}$, and vice-versa (given a

In other words, (1) a universe $\cU$ is a type including the \textit{codes or names} of types.
(2) a code (term) $\goodTerm{\Gamma}{T}{\cU}$ is not allowed to locate at type position
(i.e., at the rightmost position of a term judgement) but $El \ T$ is (i.e. we distinguish the code and the type). (3)
"c" is the ``inverse'' of $El$, obtaining  the \underline{"c"}ode of a
type, with the following \textit{two (omitted) equations representing their semantic} $"c" (El~T) \equiv T$ and $El
("c"~T) \equiv T$---notice how the two sides of each equality is term of the same type thus well-typed. More concretely,

@@

\underline{\textit{\textbf{Substitution Calculus.}}}
We quickly review the substitution calculus.
Substitution is very (if not the most) important because (dependent) functions are
first-class citizens, and lambdas can wrap (almost) every parts and can return
(almost) everything, thus we require \textbf{the resulting substitution to
commute} with most (if not all) type/term constructs.

@@

\subsubsection{\textbf{Core Syntax of the Extended MLTT}}
The core language builds on MLTT and adds four new judgement forms.
Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is just a
shorthand for term judgement
$\goodTerm{\Gamma , \cC \sigma}{\_}{A[\pi_1]}$ where $A$ is a type.

\begin{figure}[!htb]
  \begin{align*}
    &\Gamma~:~"Con"&&\mathboxtext{\goodCtx{\Gamma}{}} &&T~:~"Ty"~\Gamma&&\mathboxtext{\goodType{\Gamma}{T}{}}  &&\gamma~:~"Tms"~\Gamma~\Delta&&\mathboxtext{\goodSub{\Gamma}{\gamma}{\Delta}}  \\ & t~:~"Tm"~\Gamma~T&&\mathboxtext{\goodTerm{\Gamma}{t}{T}} 
    &&\sigma~:~"Sig"~\Gamma~n&&\mathboxtext{\goodSig{\Gamma}{\sigma}{n}}  &&\tau~:~"WSig"~\Gamma~n&&\mathboxtext{\goodWSig{\Gamma}{\tau}{n}} \\ &h~:~"Inh"~\Gamma~\sigma_1~\sigma_2&&\mathboxtext{\goodInh{\Gamma}{h}{\sigma}{\tau}} && f~:~"Seal"~\Gamma~\sigma~T&&\mathboxtext{\goodSeal{\Gamma}{f}{\sigma}{A}}\YZ{A little unfriendly to the reader that there are three different things that sigma can range over. Besides, later you also use gamma to range over substitutions and s to range over W-types, which add to the confusion.}\EDJreply{Fair point. I have fixed fix sigma and tau to be about signatures and W signature, and use gamma as substitutions}
  \end{align*}
\caption{Judgement forms, as Agda types and as presented in this section.}
\end{figure}

\begin{figure}[!htb]
  \begin{minipage}[b]{0.3\linewidth}
      $$
      \Rule[name=Tm]
      {\goodCtx{\Gamma}{i} \quad \goodType{\Gamma}{T}{j}}
      {\goodTerm{\Gamma}{\_}{T}}
      $$
      $$
      \Rule[name=Tms]
      {\goodCtx{\Gamma}{i} \quad \goodCtx{\Delta}{i}}
      {\goodSub{\Gamma}{\_}{\Delta} \text{ is well-formed}}
      $$
      $$
      \Rule[name=Sig]
      {\goodCtx{\Gamma}{i} \quad n \in \nat}
      {\goodSig{\Gamma}{\_}{n} \text{ is well-formed}}
      $$
      $$
      \Rule[name=Seal]
      {\goodCtx{\Gamma}{i} \quad \goodSig{\Gamma}{\sigma}{n} 
      \quad \goodType{\Gamma}{A}{} }
      {\goodSeal{\Gamma}{\_}{\sigma}{A} \text{ is well-formed}}
      $$\YZ{Doesn't the left column show well-formedness rules for ONLY TWO of the four new judgments? Besides, are these rules used anywhere in the following? Seems a bit unusual to have well-formedness rules for judgments per se.}\EDJreply{Yes. Right column show all the well-formedness rule for all the judgement. That is unusual because it is a direct translation from QIIT. }
  \end{minipage}
  \begin{minipage}[b]{0.6\linewidth}
    \begin{minted}[]{agda}
      data Con   : Set 
      data Ty    : Con → Set   
      data Tms   : Con → Con → Set 
      data Tm    : (Γ : Con) → Ty Γ  → Set 
      data Sig  : Con → ℕ → Set
      data WSig : Con → ℕ → Set 
      data Inh  : (Γ : Con) → Sig Γ n → Sig Γ m → Set
      Seal : (Γ : Con) → Sig Γ n → Sig Γ n → Set
    \end{minted}
  \end{minipage}

\caption{Well-formedness rules of the judgements}
\end{figure}

>>>>

\subsubsection{\textbf{Warm Up: Review Declarative MLTT}}

We highlight some specialities of declarative style in QIIT that differ from a
conventional setting: 
(1) This declarative formulation of type theory does not
have an operational semantics but only judgemental equality (e.g. $\beta$-reduction is now $\beta$-equality).\footnote{Later we will prove
canonicity which can imply a certain level of computational ability of
our theory} 
(2) We directly use the equality from QIIT to represent judgemental equality in our syntax, relieve from us the burden to explicit mentioning equivalence closure; however, this also mean that, \textit{every piece of the syntax data}, including all kinds of judgements (QIIT type) and their derivations will have equality information upon them  
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
% (3) We are dealing with quotient data, so we are actually manipulating
% \textit{equivalence classes of terms} instead of one representative (term) from the equivalence class.
(3) Instead of using meta-level substitution, we use explicit
substitution and de Bruijn indices in our intrinsic formulation. This is
also called \textit{substitution calculus} in the literature and is favoured due
to its closedness to categorical semantic. We will go through these details one-by-one.

\begin{figure}[!htb]
  \begin{align*}
    &\Gamma~:~"Con"&&\mathboxtext{\goodCtx{\Gamma}{}} &&T~:~"Ty"~\Gamma&&\mathboxtext{\goodType{\Gamma}{T}{}}  &&\gamma~:~"Tms"~\Gamma~\Delta&&\mathboxtext{\goodSub{\Gamma}{\gamma}{\Delta}}  && t~:~"Tm"~\Gamma~T&&\mathboxtext{\goodTerm{\Gamma}{t}{T}} 
  \end{align*}
\caption{Judgement forms, as Agda types and as presented in this section.}\label{fig:vanillaMLTTjudgeQIIT}
\end{figure}
\begin{figure}[!htb]
  \begin{minipage}[b]{0.3\linewidth}
      $$
      \Rule[name=Tm]
      {\goodCtx{\Gamma}{i} \quad \goodType{\Gamma}{T}{j}}
      {\goodTerm{\Gamma}{\_}{T}}
      $$
      $$
      \Rule[name=Tms]
      {\goodCtx{\Gamma}{i} \quad \goodCtx{\Delta}{i}}
      {\goodSub{\Gamma}{\_}{\Delta} \text{ is well-formed}}
      $$
  \end{minipage}
  \begin{minipage}[b]{0.6\linewidth}
    \begin{minted}[]{agda}
      data Con   : Set 
      data Ty    : Con → Set   
      data Tms   : Con → Con → Set 
      data Tm    : (Γ : Con) → Ty Γ  → Set 
    \end{minted}
  \end{minipage}

\caption{Well-formedness rules of the judgements}\label{fig:vanillaMLTTjudgewellformed}
\end{figure}
\underline{\textit{\textbf{Judgements and Universe.}}}
% Judgements in QIIT
We formalize our language as a deep embedding in a type theory using QIIT types. Every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms from (vanilla) MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. The QIITs corresponding of these four judgements are
mutually recursively defined together, and declared in \cref{fig:vanillaMLTTjudgeQIIT}. Our presentation
follows the style of~\citet{kaposi2019gluing} and thus a predicative MLTT, but for readability we
omit the universe levels.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the (pseudo) Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, in \cref{fig:vanillaMLTTjudgewellformed}, we spell out these \textit{well-formedness rules} of type
judgements and their (pseudo-)Agda counterparts to make things more accessible. In (pseudo-)Agda, well-formedness rules are just well-typedness of (indexed) QIIT family---(the judgement)"Tm~$\Gamma$~$T$" is well-typed (well-formed) only when $\Gamma$ and $T$ are both well-typed (well-formed). We omitted a bunch for latex version. Of
course, a well-formed judgement is not necessarily derivable, just as
a well-formed QIIT type is not necessarily inhabited. 
These well-formedness rules are direct translations of the \textit{well-typedness} of QIIT.

@@

% Example : how judgemental equality is used for the universe
We review the concept of ``Universe'' and at the same time showing how equality is representing semantic. In dependent type setting, function now can return type, of which the type is ``type of type'', i.e., \textit{the universe} $\cU$. We follow the style of \textit{type universe à la Tarski}~\cite{hofmann1997syntax} and
distinguish types from their \textit{codes or names} as showin in \ruleref{Code/Get Type}: given a \textit{code} (term) $\goodTerm{\Gamma}{T}{\cU}$ we can retrieve the corresponding type
$\goodType{\Gamma}{El~T}{}$, and vice-versa.

In other words, (1) a universe $\cU$ is a type including the \textit{codes or names} of types.
(2) a code (term) $\goodTerm{\Gamma}{T}{\cU}$ is not allowed to locate at type position
(i.e., at the rightmost position of a term judgement) but $El \ T$ is (i.e. we distinguish the code and the type). (3)
"c" is the ``inverse'' of $El$, obtaining  the \underline{"c"}ode of a
type, with the following \textit{two (omitted) equations representing their semantic} $"c" (El~T) \equiv T$ and $El
("c"~T) \equiv T$---notice how the two sides of each equality is term of the same type thus a well-typed equality. More concretely,

@@

\underline{\textit{\textbf{Substitution Calculus.}}}
Now we review the substitution calculus.
Substitution is very (if not the most) important because (dependent) functions are
first-class citizens, and lambdas can wrap (almost) every parts and can return
(almost) everything---once we apply the functions,  \textbf{the substitution will happen and needs to
commute} with most all type/term constructs in the function body.

@@

\subsubsection{\textbf{Core Syntax of the Extended MLTT}}
\begin{figure}[!htb]
  \begin{align*}
&\sigma~:~"Sig"~\Gamma~n&&\mathboxtext{\goodSig{\Gamma}{\sigma}{n}}  &&\tau~:~"WSig"~\Gamma~n&&\mathboxtext{\goodWSig{\Gamma}{\tau}{n}} \\ &h~:~"Inh"~\Gamma~\sigma_1~\sigma_2&&\mathboxtext{\goodInh{\Gamma}{h}{\sigma}{\tau}} && f~:~"Seal"~\Gamma~\sigma~T&&\mathboxtext{\goodSeal{\Gamma}{f}{\sigma}{A}}\YZ{A little unfriendly to the reader that there are three different things that sigma can range over. Besides, later you also use gamma to range over substitutions and s to range over W-types, which add to the confusion.}\EDJreply{Fair point. I have fixed fix sigma and tau to be about signatures and W signature, and use gamma as substitutions}
  \end{align*}
\caption{Judgement forms, as Agda types, continued}
\end{figure}
\begin{figure}[!htb]
  \begin{minipage}[b]{0.3\linewidth}
      $$
      \Rule[name=Sig]
      {\goodCtx{\Gamma}{i} \quad n \in \nat}
      {\goodSig{\Gamma}{\_}{n} \text{ is well-formed}}
      $$
      $$
      \Rule[name=Seal]
      {\goodCtx{\Gamma}{i} \quad \goodSig{\Gamma}{\sigma}{n} 
      \quad \goodType{\Gamma}{A}{} }
      {\goodSeal{\Gamma}{\_}{\sigma}{A} \text{ is well-formed}}
      $$\YZ{Doesn't the left column show well-formedness rules for ONLY TWO of the four new judgments? Besides, are these rules used anywhere in the following? Seems a bit unusual to have well-formedness rules for judgments per se.}\EDJreply{Yes. Right column show all the well-formedness rule for all the judgement. That is unusual because it is a direct translation from QIIT. }
  \end{minipage}
  \begin{minipage}[b]{0.6\linewidth}
    \begin{minted}[]{agda}
      data Sig  : Con → ℕ → Set
      data WSig : Con → ℕ → Set 
      data Inh  : (Γ : Con) → Sig Γ n → Sig Γ m → Set
      Seal : (Γ : Con) → Sig Γ n → Sig Γ n → Set
    \end{minted}
  \end{minipage}

\caption{Well-formedness rules of the extended judgements}
\end{figure}

The core language builds on MLTT and adds four new judgement forms.
Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is just a
shorthand for term judgement
$\goodTerm{\Gamma , \cC \sigma}{\_}{A[\pi_1]}$ where $A$ is a type.
